`{peacesciencer}`: An R Package for Quantitative Peace Science Research

### 2021-12-23 user-guide/ms.Rmd
update

Finally, the creation and presentation of data in peace science should be 100% robust and transparent, which `{peacesciencer}` takes seriously in the following ways. The website for `{peacesciencer}` has several vignettes that describe its processes in some detail. These include how it provides reasonable estimates of democracy that may not be available in the Polity data or the Varieties of Democracy data and how a researcher can whittle dyadic dispute-years into true dyad-years through reasonable case exclusions. `{peacesciencer}` subjects itself to a battery of tests before publishing updates, making sure new features do not create duplicate entries in the original data (which is the surest sign of a botched merge).  The project's Github contains a publicly available `data-raw` directory that shows how every data set included (and processed) in `{peacesciencer}` was created. The function manuals included `{peacesciencer}` contain ample documentation that clarifies what the function is doing, what it returns to the user, and why it is doing it this way. Researchers can also use the project's Github to point out bugs, ask for further clarification, and propose additions. `{peacesciencer}` takes seriously the Data Access and Research Transparency Initiative (DA-RT) initiative across all political science and endeavors for maximum transparency, leveraging open source and version control software to inform users of what data it uses and how it uses the data.

@@

`{peacesciencer}` comes with a fully developed suite of built-in functions for generating some of the most widespread forms of peace science data and populating the data with important variables that recur in many quantitative analyses. The core functionality of `{peacesciencer}` reduces to two broad categories of functions. These categories are functions that create the base data of interest to a researcher and functions, called after the base data are created, that add variables of interest to the data frame or subset the base data to a handful of rows that the researcher deems appropriate for analysis.[^leftjoin] Table \@ref(tab:tab-createfunctions) and Table \@ref(tab:tab-addfunctions) list these core functions as of version 0.7, the version of this package slated for release alongside the manuscript.[^helperfunctions]

@@

Table \@ref(tab:tab-createfunctions) lists functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 0.7, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. dyadic-dispute-year, leader-months, state-quarters).

@@

`{peacesciencer}`'s coverage focuses mostly on data that are released as standalone data sets for download, especially those in the Correlates of War or Gleditsch-Ward ecosystem of data. Data that can be obtained from a stable advanced programming interface---like the World Bank, for example---can be obtained through those other means [e.g. @arelbundock2021w]. Its coverage will assuredly expand with new additions of interest to the peace science community, though the package already offers a lot to meet researcher needs.

@@

For example, assume the researcher wants just all politically relevant, non-directed dyad-years, where political relevance is traditionally understood as a dyadic relationship involving some form of a contiguity relationship or a major power [@lemkereed2001rprd]. `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))` created the full universe of non-directed dyad-years from 1816 to 2010, though this full universe includes "irrelevant" dyads like Nigeria-Mongolia and Canada-Estonia. Reducing the data to just politically relevant dyads is simple in `{peacesciencer}` and `{tidyverse}`. Users first create the data they want (here: `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))`), follow it with the pipe operator (`%>%`), and then add another function from `{peacesciencer}` (here: `filter_prd()`, which also quietly executes `add_contiguity()` and `add_cow_majors()`).

@@

If the user wants to move these data into Stata for analysis, they can save it to their current working directory with a command like `haven::write_dta(Data, "my-data.dta")` and import it into Stata when they are done. No matter, `{peacesciencer}` has pre-processed, cleaned, recoded, and merged the desired data that has greatly reduced the time and energy a researcher might otherwise spend doing something like hard-coding `-9s` in these data to be `NA` in the capabilities data. In this particular application, it has already created the main data required for a replication of @bremer1992ddc. There is only some slight data work to create the desired indicators for a statistical model of conflict onset, like a dummy variable for land-contiguity, the presence of a major power in the dyad, and some "weak-link" indicators of militarization, relative power in the dyad, level of democracy in the dyad (using the @marquez2016qme for extending the Unified Democracy Scores data), and the GDP per capita in the dyad. Table \@ref(tab:tab-dd) is a formatted version of the results of a logistic regression model of conflict onset using these "dangerous dyads" indicators. Users typically do not end their analysis here---often looking for new predictors of conflict onset with these covariates in mind---but `{peacesciencer}` greatly reduces the time and energy researchers must invest into cleaning and processing data for analysis.

@@

`{peacesciencer}` is capable of generating data for replications of analyses at multiple levels beyond dyad-year. Suppose a researcher wants to create a state-year data frame to conduct an analysis of civil conflict onset analogous to Fearon and Laitin's [-@fearonlaitin2003eicw] well-cited analysis of civil conflict onset, but using UCDP conflict data and the Gleditsch-Ward state system for creating the appropriate universe of state-years. The pipe-based workflow will start with `create_stateyears(system = 'gw', subset_years = c(1946:2019))`, creating the full universe of Gleditsch-Ward state years and subsetting them to just 1946-2019 (because the UCDP data included in `{peacesciencer}` include just the observations in that time frame). Next, we can use the `add_ucdp_acd()` function to return information about ongoing UCDP conflicts and onsets for these states. `add_ucdp_acd()` takes three arguments: `type`, `issue`, and `only_wars`. `type` is an optional argument for the type of armed conflicts for which the researcher wants information. Options include "extrasystemic", "interstate", "intrastate", and "II" (short for "internationalized intrastate"). If no `type` is specified, the function returns information about ongoing disputes and onsets for all states for all tyeps of conflict. If the user wants information about multiple types of conflict--say: intra-state wars and internationalized intra-state wars---they can specify that as a character vector (e.g. `type = c("intrastate", "II")`). `issue` is another optional argument for what issue types of conflicts the user wants beyond the type of armed conflict. Options include "territory", "government", and "both". If no `issue` is specified, the function returns information for all conflicts regardless of the particular issue. `only_wars` is an argument that subsets the data to just those with the intensity levels of "war" when `only_wars = TRUE`. The argument defaults to `FALSE`, returning information about conflicts with at least 25 deaths in addition to the conflicts with more than 1,000 deaths. In this application, `add_ucdp_acd(type = "intrastate", only_wars = FALSE))` returns state-year information about ongoing intra-state conflicts over any issue and at either of UCDP's severity thresholds.[^adducdpacdalso]

>>>>

Finally, the creation and presentation of data in peace science should be 100% robust and transparent, which `{peacesciencer}` takes seriously in the following ways. The website for `{peacesciencer}` has several vignettes that describe its processes in some detail. These include how it provides reasonable estimates of democracy that may not be available in the Polity data or the Varieties of Democracy data and how a researcher can whittle dyadic dispute-years into true dyad-years through reasonable case exclusions. `{peacesciencer}` subjects itself to a battery of tests before publishing updates, making sure new features do not create duplicate entries in the original data (which is the surest sign of a botched merge).  The project's Github contains a publicly available `data-raw` directory that shows how every data set included (and processed) in `{peacesciencer}` was created. The function manuals included `{peacesciencer}` contain ample documentation that clarify what each function is doing, what it returns to the user, and why it is doing it this way. Researchers can also use the project's Github to point out bugs, ask for further clarification, and propose additions. `{peacesciencer}` takes seriously the Data Access and Research Transparency Initiative (DA-RT) initiative across all political science and endeavors for maximum transparency, leveraging open source and version control software to inform users of what data it uses and how it uses the data.

@@

`{peacesciencer}` comes with a fully developed suite of built-in functions for generating some of the most widespread forms of peace science data and populating the data with important variables that recur in many quantitative analyses. The core functionality of `{peacesciencer}` reduces to two broad categories of functions. These categories are functions that create the base data of interest to a researcher and functions, called after the base data are created, that add variables of interest to the data frame or subset the base data to a handful of rows that the researcher deems appropriate for analysis.[^leftjoin] Table \@ref(tab:tab-createfunctions) and Table \@ref(tab:tab-addfunctions) list these core functions as of version 1.0, the version of this package slated for release alongside the manuscript.[^helperfunctions]

@@

Table \@ref(tab:tab-createfunctions) lists functions that create base data frames for a researcher starting an original project. They serve as functions that communicate the units of analysis supported in this package and that the package is capable of generating for an interested user. For example, `create_stateyears()` will generate the full universe of state-years from the Correlates of War [@cowstates2016, v. 2016] or Gleditsch-Ward [@gledtischward1999rlis, v.2017] system, encompassing all state-years to the most recently concluded calendar year, depending on the arguments supplied to the user in the function. `create_leaderyears()` will generate the full universe of leader-years from the Archigos leader data [@goemansetal2009ia, v. 4.1], optionally standardizing leader-years to the Gleditsch-Ward or Correlates of War state system data. As of version 1.0, `{peacesciencer}` is capable of creating full dyad-year data, leader-day data, leader-dyad-year data, leader-year data, state-day data, and state-year data.[^leaderdyadyear] A vignette on the package's website shows how users can create other forms of data from these functions as well (e.g. dyadic-dispute-year, leader-months, state-quarters).

@@

`{peacesciencer}`'s coverage focuses mostly on data that are released as standalone data sets for download, especially those in the Correlates of War or Gleditsch-Ward ecosystem of data. Data that can be obtained from a stable advanced programming interface---like the World Bank, for example---can be obtained through those other means [e.g. @arelbundock2021w]. Its coverage will assuredly expand with new additions of interest to the peace science community, though the package already offers a lot to meet researcher needs.[^ps_version]

[^ps_version]: The version of `{peacesciencer}` slated for release includes the most up-to-date-possible versions of data sets referenced in this manuscript. One helper function, `ps_version()`, will inform the user about the versions of the data included in the installed package. `{peacesciencer}` promises prompt package updates as new versions become available.

@@

For example, assume the researcher wants just all politically relevant, non-directed dyad-years, where political relevance is traditionally understood as a dyadic relationship involving some form of a contiguity relationship or a major power [@lemkereed2001rprd]. `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))` created the full universe of non-directed dyad-years from 1816 to 2010, though this full universe includes "irrelevant" dyads like Nigeria-Mongolia and Estonia-Rwanda. Reducing the data to just politically relevant dyads is simple in `{peacesciencer}` and `{tidyverse}`. Users first create the data they want (here: `create_dyadyears(directed = FALSE, subset_years = c(1816:2010))`), follow it with the pipe operator (`%>%`), and then add another function from `{peacesciencer}` (here: `filter_prd()`, which also quietly executes `add_contiguity()` and `add_cow_majors()`).

@@

If the user wants to move these data into Stata for analysis, they can save it to their current working directory with a command like `haven::write_dta(Data, "my-data.dta")` and import it into Stata when they are done. No matter, `{peacesciencer}` has pre-processed, cleaned, recoded, and merged the desired data that has greatly reduced the time and energy a researcher might otherwise spend doing something like hard-coding `-9s` in these data to be `NA` in the capabilities data. In this particular application, it has already created the main data required for a replication of @bremer1992ddc. There is only some slight data work to create the desired indicators for a statistical model of conflict onset, like a dummy variable for land-contiguity, the presence of a major power in the dyad, and some "weak-link" indicators of militarization, relative power in the dyad, level of democracy in the dyad (using the @marquez2016qme method for extending the Unified Democracy Scores data), and the GDP per capita in the dyad. Table \@ref(tab:tab-dd) is a formatted version of the results of a logistic regression model of conflict onset using these "dangerous dyads" indicators. Users typically do not end their analysis here---often looking for new predictors of conflict onset with these covariates in mind---but `{peacesciencer}` greatly reduces the time and energy researchers must invest into cleaning and processing data for analysis.

@@

`{peacesciencer}` is capable of generating data for replications of analyses at multiple levels beyond dyad-year. Suppose a researcher wants to create a state-year data frame to conduct an analysis of civil conflict onset analogous to Fearon and Laitin's [-@fearonlaitin2003eicw] well-cited analysis of civil conflict onset, but using UCDP conflict data and the Gleditsch-Ward state system for creating the appropriate universe of state-years. The pipe-based workflow will start with `create_stateyears(system = 'gw', subset_years = c(1946:2019))`, creating the full universe of Gleditsch-Ward state years and subsetting them to just 1946-2019 (because the UCDP data included in `{peacesciencer}` include just the observations in that time frame). Next, we can use the `add_ucdp_acd()` function to return information about ongoing UCDP conflicts and onsets for these states. `add_ucdp_acd()` takes three arguments: `type`, `issue`, and `only_wars`. `type` is an optional argument for the type of armed conflicts for which the researcher wants information. Options include "extrasystemic", "interstate", "intrastate", and "II" (short for "internationalized intrastate"). If no `type` is specified, the function returns information about ongoing disputes and onsets for all states for all types of conflict. If the user wants information about multiple types of conflict--say: intra-state wars and internationalized intra-state wars---they can specify that as a character vector (e.g. `type = c("intrastate", "II")`). `issue` is another optional argument for what issue types of conflicts the user wants beyond the type of armed conflict. Options include "territory", "government", and "both". If no `issue` is specified, the function returns information for all conflicts regardless of the particular issue. `only_wars` is an argument that subsets the data to just those with the intensity levels of "war" when `only_wars = TRUE`. The argument defaults to `FALSE`, returning information about conflicts with at least 25 deaths in addition to the conflicts with more than 1,000 deaths. In this application, `add_ucdp_acd(type = "intrastate", only_wars = FALSE))` returns state-year information about ongoing intra-state conflicts over any issue and at either of UCDP's severity thresholds.[^adducdpacdalso]
